---
title: "Trout Creek Watershed"
author: "Colleen Hoskins"
format: 
  html:
    code-fold: true
---

## **Introduction**

Trout Creek was historically one of the most productive fisheries in the Tahoe Basin. In 1859, logging began in the Trout Creek Area. During the logging period, 13 miles of railroad, 16 miles of wagon haul roads, 2 miles of “V” flume and 28 logging camps were constructed in the Trout Creek watershed, causing erosion along Trout Creek and Saxton Creek (TRCD, 2013). Post-logging, the Trout Creek watershed was used to graze cattle and sheep. During this time, a ditch system and a series of small dams were developed along the tributaries of Trout Creek and Cold Creek (TRCD, 2013). Excessive grazing degraded watershed conditions. In 1996, the dams were removed from Cold Creek (California Tahoe Conservancy, 2022). Since 1996, projects within the Trout Creek watershed were conducted to restore the watershed to its historic condition. In 2021, the Caldor Fire burned areas within the Trout Creek Watershed, affecting the watershed’s streams (California Tahoe Conservancy, 2022). Moving forward, plans are being developed for restoration projects in several sections of Trout Creek and Cold Creek.  

This analysis compares flow and components of the water budget for the pre-restoration period before 1996 and the restoration period from 1996 onward.

```{r}
#| warning: false
#| message: false
#| echo: false 
#| results: "hide"

source("data/setup.R")

tc_watershed <- st_read("tc_shapefile/layers/globalwatershed.shp")
tc_point <- st_read("tc_point/trout_creek_gauge_1.shp")
tc_point$site_id <- "USGS-10336610"
```

```{r}
#| fig-cap: "Figure 1: Map of Trout Creek watershed boundary and location of stream gauge USGS-10336790."
#| warning: false
#| echo: false
#| messages: false

tmap_mode("view")

tm_shape(tc_watershed) +
  tm_borders(col = "blue") +
  tm_shape(tc_point) +
  tm_bubbles(col = "red") +  
  tm_basemap("Esri.WorldStreetMap")
```

## **Watershed Characteristics**

The Trout Creek Watershed is located on the southern side of Lake Tahoe, California, adjacent to the Nevada state border (Figure 1). The main water features include Saxton Creek, Cold Creek, Trout Creek and Heavenly Valley Creek. These features converge into the main channel of Trout Creek, which flows north before draining into Lake Tahoe. The drainage area above the stream gauge is 36.7 square miles (Table 1).

The highest elevation of Trout Creek Watershed Is 10878 feet and the minimum elevation at the stream gauge is 6261 feet. The mean slope of the watershed is 30.1%, with steep, mountainous terrain in the southern region that becomes flatter toward the northern part as it nears the lake. The largest listed land cover type in the area is forest, which covers 43.1% of the watershed (Table 1). Urban land covers 3.1%, and lakes and ponds cover less than 1% of the area. Other land cover types are not specified by Web Soil Survey. Soils in the watershed are dominantly composed of various types of sand, including loamy, cobbly, coarse, gravelly, and stony sands, along with combinations of these textures (Table 2). They are primarily classified as excessively well-drained.

| Parameter Description                   | Value     |
|:----------------------------------------|:----------|
| drainage area                           | 36.7 mi²  |
| mean basin slope computed from 30 m DEM | 30.1 %    |
| mean basin elevation                    | 7932 ft   |
| maximum basin elevation                 | 10878 ft  |
| minimum basin elevation                 | 6261 ft   |
| area covered by forest                  | 43.1 %    |
| area covered by lakes and ponds         | 0.08881 % |
| area of developed (urban) land          | 3.1 %     |

: Table 1: Characteristics of the Trout Creek watershed.

+-------------------------------------+-------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------+-----------------------------------+
| Soil Series                         | Primary Texture                                                                                                         | Drainage Class                                    | Ksat                              |
+=====================================+=========================================================================================================================+===================================================+===================================+
| Cassenai                            | gravelly loamy coarse sand, cobbly loamy coarse sand, gravelly coarse sand                                              | somewhat excessively drained                      | high                              |
|                                     |                                                                                                                         |                                                   |                                   |
|                                     |                                                                                                                         |                                                   | (1.98 to 5.95 in/hr)              |
+-------------------------------------+-------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------+-----------------------------------+
| Temo-Witefels complex               | gravelly loamy coarse sand                                                                                              | excessively drained/ somewhat excessively drained | moderately low to moderately high |
|                                     |                                                                                                                         |                                                   |                                   |
|                                     |                                                                                                                         |                                                   | (0.01 to 1.42 in/hr)              |
+-------------------------------------+-------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------+-----------------------------------+
| Dagget                              | very gravelly loamy coarse sand                                                                                         | excessively drained                               | moderately low to moderately high |
|                                     |                                                                                                                         |                                                   |                                   |
|                                     |                                                                                                                         |                                                   | (0.01 to 1.42 in/hr)              |
+-------------------------------------+-------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------+-----------------------------------+
| Whitell-Jobsis-Rock outcrop complex | very cobbly loamy coarse sand, gravelly loamy coarse sand, very gravelly coarse sand, extremely stony loamy coarse sand | somewhat excessively drained                      | moderately low to moderately high |
|                                     |                                                                                                                         |                                                   |                                   |
|                                     |                                                                                                                         |                                                   | (0.01 to 1.42 in/hr)              |
+-------------------------------------+-------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------+-----------------------------------+

: Table 2: Characteristics of major soil types of the Trout Creek watershed.

## Watershed Analysis

Discharge data for USGS-10336780, the Trout Creek stream gauge, spans water years 1961-2023. Precipitation, grass reference evapotranspiration, snow water equivalent, and soil moisture at a depth of 10-40 cm were downloaded from climateengine.org for water years 1981-2023. Precipitation and grass reference evapotranspiration trends are calculated from the dataset GridMET - 4km -daily. Snow water equivelent and soil moisture trends are calculated from the dataset WILDAS - 1.1 km - daily.

Discharge minimum, maximum, mean, and standard deviation are all higher in the restoration group than in the pre-restoration group, indicating greater discharge variability (Table 3). The restoration period saw the highest daily discharge on January 2, 1997, at 12.80 mm (Figure 2). The highest annual discharge occurred in 2017, totaling 152.80 mm. Discharge exceedance probability (p) shows slightly higher low and high flows in the restoration group compared to the pre-restoration group (Figures 3 & 4).

The watershed is semi-arid with an average annual precipitation of 895.91 mm and an aridity index (P/ET) of 0.79. The runoff ratio (Q/P) is 0.39, meaning that over half of precipitation does not become runoff. This could be due to evapotranspiration and the watershed's excessively drained soils.

Mean annual precipitation increased slightly from 888.26 mm pre-restoration to 899.96 mm in the restoration period. Mean annual discharge also increased slightly from 337.80 mm pre-restoration to 357.09 mm post-restoration. Mean annual evapotranspiration increased from 1083.18 mm pre-restoration to 1153.28 mm post-restoration. In contrast, both mean annual snow water equivalent and soil moisture decreased from the pre-restoration to the restoration group. Snow water equivalent dropped from 311.69 mm to 297.45 mm, and soil moisture dropped from 50.04 mm to 48.83 mm. Annual trends show that evapotranspiration increased, while soil moisture decreased from 1981 to 2023 (Figure 5).

Three notable years occurred: 1983, 2017, and 2021. In 1983, the watershed recorded the highest annual average soil moisture (54.59 mm) and the lowest annual average evapotranspiration (941.37 mm). In 2017, the watershed experienced the highest recorded annual discharge (1052.80 mm) and precipitation (1888.09 mm). In 2021, the watershed recorded the lowest annual soil moisture (42.91 mm) and the highest annual evapotranspiration (1250.37 mm). The Caldor Fire occurred during water years 2021 and 2022. Specifically, the Caldor Fire started on 8/14/2021 and was fully contained on 10/25/2021. Rain and snowmelt inputs were low from June to September of 2021, causing soil moisture to decline during that period (Figure 7). A precipitation event in mid-October of 2021 increased soil moisture and coincided with the containment of the fire.

Monthly trends show that precipitation is elevated from October through April, with a decline from May through September (Figure 6). Snow accumulates during winter, with snowmelt typically occurring between April and July. Soil moisture decreases during the spring and summer months, continuing through August and September, before increasing with higher precipitation in winter. Discharge peaks from March through August, typically in May or June. Evapotranspiration is lowest in December and January, gradually increasing and peaking in July, before gradually declining. Evapotranspiration rises in warmer months and declines in the winter months.

|                 |           |           |                        |
|-----------------|-----------|-----------|------------------------|
| **Period**      | **Max Q** | **Min Q** | **Standard Deviation** |
| Pre restoration | 801.80    | 95.88     | 182.50                 |
| Restoration     | 1052.80   | 108.22    | 240.12                 |

: Table 3: Annual mean, maximum, minimum, and standard deviation for discharge (Q) of Trout Creek for the pre-restoration period (water years 1961-1995) and restoration period (1996-2023).

## Figures

```{r}
#| fig-cap: "Figure 2: Annual, daily, and total sum discharge for Trout Creek for water years 1960-2023."
#| warning: false

# Create a table for the stream gauge
trout_creek_table <- tibble(site_no = '10336780', gauge = 'Trout Creek')

# Get data from USGS
trout_creek <- readNWISdv(
  siteNumber = trout_creek_table$site_no,
  parameterCd = '00060',
  startDate = "1960-10-01",
  endDate = "2024-09-30"
) %>%
  renameNWISColumns() %>%
  left_join(trout_creek_table, by = 'site_no') %>%
  select(-'Flow_cd')

# Convert discharge from cfs to mm/day (drainage area is 36.7 square miles)
trout_creek <- trout_creek %>%
  mutate(Q_mm = Flow * 86400 / 36.7 / 5280 / 5280 * 304.8)

# Calculate water years
trout_creek <- trout_creek %>%
  mutate(
    month = month(Date),
    wyear = if_else(month > 9, year(Date) + 1, year(Date))
  )

# Separate into periods to analyze: the pre-restoration period is before 1996.
# The restoration period is equal to and following 1996.
# Calculate annual totals for Q
Q_annual <- trout_creek %>%
  mutate(group = ifelse(wyear < 1996, "pre-restoration", "restoration")) %>%
  group_by(group, wyear) %>%
  summarize(Q_mm = sum(Q_mm), count = n())

# Graph annual discharge with a scatter plot
Q <- Q_annual %>%
  ggplot(mapping = aes(x = wyear, y = Q_mm, color = group)) +
  geom_point() +
  theme_bw() + 
  labs(x = "Water Year", y = "Annual Q (mm)", color = "Group") +
  guides(color = guide_legend(title = NULL))

# Graph annual discharge with a box plot
Q_box <- Q_annual %>%
  ggplot(mapping = aes(x = group, y = Q_mm, fill = group)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "", y = 'Q (mm)') +
  guides(fill = guide_legend(title = NULL))

# Graph annual discharge with a violin plot
Q_violin <- Q_annual %>%
  ggplot(mapping = aes(x = group, y = Q_mm, fill = group)) +
  geom_violin() +
  theme_bw() +
  labs(x = "", y = 'Q (mm)') +
  guides(fill = guide_legend(title = NULL))

# Plot daily discharge
Q_daily <- trout_creek %>%
  mutate(group = ifelse(wyear < 1996, "pre-restoration", "restoration")) %>%
  ggplot(mapping = aes(x = wyear, y = Q_mm, color = group)) +
  geom_line() +
  theme_bw() + 
  labs(x = "Water Year", y = "Daily Q (mm)", color = "Group") +
  guides(color = guide_legend(title = NULL))

# Stitch plots together
ggarrange(Q_violin, Q_box, Q_daily, Q, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")
```

```{r}
#| fig-cap: "Figure 3: Discharge(Q) exceedance probability (p) for Trout Creek watershed for water years 1961-2023."
#| warning: false

#Calculate flow duration curve:
#exceedance probability (p) = rank/n+1
pre_restoration <- trout_creek %>%
  filter(wyear < 1996) %>%
  mutate(
    group='pre-restoration',
    rank=rank(Flow),
    p=rank/12784 
  )

restoration <- trout_creek %>%
  filter(wyear >= 1996) %>%
  mutate(
    group='restoration',
    rank=rank(Flow),
    p=rank/10594 
  )

fdc=rbind(pre_restoration,restoration)

#Plot fcd
fdc_plot <- fdc %>%
  ggplot(mapping=aes(x=p,y=Flow,color=group, linetype=group))+
  geom_line()+
  theme_bw()+
  labs(x = "Exceedance Probability (p)",y = "Flow (cfs)") +
  guides(color = guide_legend(title = NULL),
         linetype = guide_legend(title = NULL))

ggplotly()
```

```{r}
#| fig-cap: "Figure 4: Discharge(Q) exceedance probability (p) at Log 10 for Trout Creek watershed for water years 1961-2023."
#| warning: false

#Plot fdc scale_y_log10() to better visualize data
#Optional: can put limits within the log; example: scale_y_log10(100,1000)
fdc_plot_log <- fdc %>%
  ggplot(mapping=aes(x=p,y=Flow,color=group, linetype=group))+
  scale_y_log10()+
  geom_line()+
  theme_bw()+
  labs(x = "Exceedance Probability (p)",y = "Flow (cfs) - Log 10") +
  guides(color = guide_legend(title = NULL),
         linetype = guide_legend(title = NULL))

ggplotly()
```

```{r}
#| fig-cap: "Figure 5: Annual sum for precipitation(P), grass reference evapotranspiration(ET), and discharge (Q). Annual peak snow water equivalent(SWE). Annual mean soil moisture(SM). Values for water years 1980-2023."
#| warning: false

# Read CSV files for P, ET, SM, and SWE:
P_TC <- read.csv('data/TC_Precip_Gridmet.csv')
ET_TC <- read.csv('data/TC_ET_Gridmet.csv')
SM_TC <- read.csv('data/TC_SM_10_40_Wildas.csv')
SWE_TC <- read.csv('data/TC_SWE_Wildas.csv')

# Filter Q_1981 based on the year range
Q_1981 <- trout_creek %>%
  filter(wyear >= 1981 & wyear <= 2023)

# Convert Date Format
P_TC$Date = mdy(P_TC$Date)

# Join P_TC with Q_1981
P_TC <- P_TC %>%
  left_join(Q_1981 %>% select(Date, Q_mm, wyear), by = "Date")

# Bind Other Variables
P_TC$ET_mm = ET_TC$ET_mm
P_TC$SM_mm = SM_TC$SM_mm
P_TC$SWE_mm = SWE_TC$SWE_mm

#Separate Pre-restoration and Restoration Periods
TC_climate <- P_TC %>%
  mutate(
    group = case_when(
      wyear < 1996 ~ "pre-restoration",
      wyear >= 1996 ~ "restoration"))


#Calculate Annual Trends
climate_annual <- suppressMessages(TC_climate %>%
  group_by(wyear,group) %>%
  summarize(P_mm=sum(P_mm),
            Q_mm=sum(Q_mm),
            ET_mm=sum(ET_mm),
            peakSWE=max(SWE_mm),
            meanSM=mean(SM_mm),
            ))

#Plot Annual Sum Q
Q_1981 <- Q_annual %>%
  ggplot(mapping=aes(x=wyear,y=Q_mm,color=group)) +
  geom_point() +
  theme_bw() + 
  labs(x = "Water Year", y = "Q (mm)", color = "Group") +
  guides(color = guide_legend(title = NULL))

#Plot Annual Sum P
P <- climate_annual %>%
  ggplot(mapping=aes(x=wyear,y=P_mm,color=group)) +
  geom_point() +
  theme_bw() + 
  labs(x = "Water Year", y = "P (mm)", color = "Group") +
  guides(color = guide_legend(title = NULL))

#Plot Annual Sum ET
ET <- climate_annual %>%
  ggplot(mapping=aes(x=wyear,y=ET_mm,color=group)) +
  geom_point() +
  theme_bw() + 
  labs(x = "Water Year", y = "ET (mm)", color = "Group") +
  guides(color = guide_legend(title = NULL))

#Plot Annual Peak SWE
SWE <- climate_annual %>%
  ggplot(mapping=aes(x=wyear,y=peakSWE,color=group)) +
  geom_point() +
  theme_bw() + 
  labs(x = "Water Year", y = "SWE (mm)", color = "Group") +
  guides(color = guide_legend(title = NULL))

#Plot Annual Mean SM
SM <- climate_annual %>%
  ggplot(mapping=aes(x=wyear,y=meanSM,color=group)) +
  geom_point() +
  theme_bw() + 
  labs(x = "Water Year", y = "SM (mm)", color = "Group") +
  guides(color = guide_legend(title = NULL))

ggarrange(P,ET,SWE,SM,Q_1981, ncol=2,nrow=3, common.legend = TRUE, legend = "bottom")
```

```{r}
#| fig-cap: "Figure 6: Monthly averages for Trout Creek watershed for water years 1961 – 2023. Mean monthly sums for precipitation(P), grass reference evapotranspiration(ET), and discharge (Q). Mean monthly snow water equivalent(SWE) and soil moisture(SM)."
#| warning: false

#Calculate monthly mean for P, Q, ET, SWE, and SM. For the flux variables P, Q and ET, calculate monthly sums before averaging: 
monthly_mean_flux <- suppressMessages(TC_climate %>%
  mutate(month=month(Date)) %>%
  group_by(month, group, wyear) %>%
  summarize(
    Q_mm_sum = sum(Q_mm),
    P_mm_sum = sum(P_mm),
    ET_mm_sum = sum(ET_mm)) %>%
   summarize(
    Q_mm = mean(Q_mm_sum),
    P_mm = mean(P_mm_sum),
    ET_mm = mean(ET_mm_sum)))

monthly_mean_stock <- suppressMessages(TC_climate %>%
  mutate(month=month(Date)) %>%
  group_by(month,group) %>%
  summarize(
    SWE_mm = mean(SWE_mm),
    SM_mm = mean(SM_mm)))

#Plot Mean Monthly SWE
SWE_mean_plot <- monthly_mean_stock %>%
  ggplot(mapping = aes(x = month, y = SWE_mm, color = group, linetype = group)) + 
  geom_line() +
  scale_x_continuous(breaks = 1:12) + 
  theme_bw() + 
  labs(x = "Month", y = "SWE (mm)", color = "Group", linetype = "Group")

#Plot Mean Monthly ET
ET_mean_plot <- monthly_mean_flux %>%
  ggplot(mapping = aes(x = month, y = ET_mm, color = group, linetype = group)) + 
  geom_line() +
  scale_x_continuous(breaks = 1:12) + 
  theme_bw() + 
  labs(x = "Month", y = "ET (mm)", color = "Group", linetype = "Group") +
  guides(color = guide_legend(title = NULL), linetype = guide_legend(title = NULL))

#Plot Mean Monthly SM
SM_mean_plot <- monthly_mean_stock %>%
  ggplot(mapping = aes(x = month, y = SM_mm, color = group, linetype = group)) + 
  geom_line() +
  scale_x_continuous(breaks = 1:12) + 
  theme_bw() + 
  labs(x = "Month", y = "SM (mm)", color = "Group", linetype = "Group") +
  guides(color = guide_legend(title = NULL), linetype = guide_legend(title = NULL))

#Plot Mean Monthly P
P_mean_plot <- monthly_mean_flux %>%
  ggplot(mapping = aes(x = month, y = P_mm, color = group, linetype = group)) + 
  geom_line() +
  scale_x_continuous(breaks = 1:12) + 
  theme_bw() + 
  labs(x = "Month", y = "P (mm)", color = "Group", linetype = "Group") +
  guides(color = guide_legend(title = NULL), linetype = guide_legend(title = NULL))

#Plot Mean Monthly Q
Q_mean_plot <- monthly_mean_flux %>%
  ggplot(mapping=aes(x=month,y=Q_mm,color=group, linetype = group)) +
  geom_line()+
  scale_x_continuous(breaks = 1:12) + 
  theme_bw() + 
  labs(x = "Month", y = "Q (mm)", color = "Group", linetype = "Group") +
  guides(color = guide_legend(title = NULL), linetype = guide_legend(title = NULL))

ggarrange(
  P_mean_plot, ET_mean_plot, Q_mean_plot, SM_mean_plot, SWE_mean_plot, 
  ncol = 2, nrow = 3,
  common.legend = TRUE,  
  legend = "bottom")
```

```{r}
#| fig-cap: "Figure 7: Rain, snowmelt, and soil moisture depth in mm for water years 2021 and 2022 (10/01/2020 – 9/30/2022) for Trout Creek watershed."
#| warning: false

#Closer Look at the Caldor Fire
caldor <- TC_climate %>%
  filter(wyear==2021 | wyear==2022) 

#Separate Snowmelt and Rain from P
caldor$rain=ifelse(caldor$SWE_mm==0,caldor$P_mm,0)
caldor$snowmelt=0

for (i in 2:730) {
  caldor[i, 'snowmelt'] <- ifelse(caldor[i, 'SWE_mm'] < caldor[i-1, 'SWE_mm'], caldor[i-1, 'SWE_mm'] - caldor[i, 'SWE_mm'], 0)
}

#Plot Rain, Snowmelt, and soil Moisture
caldor_plot <- caldor %>%
  ggplot() +
  geom_line(mapping = aes(x = Date, y = SM_mm, color = "SM (mm)")) +   
  geom_line(mapping = aes(x = Date, y = rain, color = "Rain (mm/day)")) +     
  geom_line(mapping = aes(x = Date, y = snowmelt, color = "Snowmelt (mm/day)")) + 
  geom_line(mapping = aes(x = Date, y = ET_mm, color = "ET (mm/day)")) +
  theme_bw()+
  labs(y = "",color = "Legend") +
  guides(color = guide_legend(title = NULL))+
  scale_x_date(
    breaks = "4 month",
    labels = scales::date_format("%b %Y"))

ggplotly()
```

## Conclusions

The restoration period shows increased variability in discharge compared to the pre-restoration period. This change appears to correlate more strongly with variations in precipitation than with the restoration efforts. Seasonal and annual trends suggest that precipitation and snowpack strongly influence streamflow. The observed increases in average precipitation, discharge and reference evapotranspiration, coupled with decreases in snow water equivalent and soil moisture suggest that climate variability may be driving hydrologic changes in the watershed. The most prominent trends from water years 1981 through 2023 are the consistent decrease in soil moisture and the increase in reference evapotranspiration, indicating increasing aridity in the watershed. Additionally, greater climatic variability is observed during the restoration period, with both higher and lower precipitation values, which contribute to increased fluctuations in snow water equivalent and discharge. These patterns may be influenced by more dramatic climate-driven weather patterns, which may amplify seasonal and annual variability in hydraulic conditions of Trout Creek watershed. Record high annual evapotranspiration and low soil moisture correlate to the Caldor fire, which burned through parts of the watershed in 2021.

## Sources

California Tahoe Conservancy. (2022). Greater Upper Truckee Watershed Restoration Accomplishments and Opportunities. <https://tahoe.ca.gov/programs/water-quality-watersheds/>

TRCD. (2013). Conservation Projects and Environmental Improvement Projects (EIPs) in the Trout-Cold Creek Community Watershed. *Tahoe Resource Conservation District.* [https://tahoercd.org](https://tahoercd.org/%20wp-content/uploads/2013/03/Conservation-Projects-Trout-and-Cold-Creek.pdf)

ClimateEngine.org. (2024). <https://www.climateengine.org>

USGS. (2024). StreamStats. <https://streamstats.usgs.gov/ss/>

USDA. (2024). Web Soil Survey. <https://websoilsurvey.sc.egov.usda.gov/App/WebSoilSurvey.aspx>
